Modelling plant mortality: integrating intra and interspecific approaches

### 2016-01-29 ms/ms.Rnw
Rewrote methods and tweaked introduction

\documentclass[a4paper,11pt]{article}
\usepackage[osf]{mathpazo}
\usepackage{ms}
\usepackage[]{natbib}
\usepackage{gensymb}
%\raggedright

@@

\usepackage[title, titletoc, toc]{appendix}

@@

<<preload, echo=FALSE>>=
n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique))
n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
@

@@

Wood density, the biomass invested per unit wood volume, is one functional trait that has been consistently shown to be negatively correlated with tropical tree mortality \citep{Chave:2009iy, Poorter:2008iu, Kraft:2010kq, Wright:2010fl}. The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to random events such as stem breakage \citep{Putz:1983wu, vanGelder:2006exa, Chave:2009iy}, embolism \citep{Hacke:2001kj, Jacobsen:2005fx} and fungal and pathogen attack \citep{Augspurger:1984wx}. As such, it is hypothesized that species with higher wood density are likely to exhibit lower baseline hazards. However, research has also shown that wood density may increase tolerance to low growth rates, and thus alter the growth-mortality relationship. For example, species with high wood density are more tolerant shade \citep{vanGelder:2006exa} and competition \citep{Kunstler:2016km}. Studies have also shown that wood density may reduce drought induced carbon starvation \citep{McDowell:2011dr} and the susceptibility of ‘stressed’ individuals to opportunistic invertebrate, pathogen or fungal attack \citep{Augspurger:1984wx, McDowell:2011dr}.

@@

In this study, we reconcile both intraspecific and interspecific theory by developing an individual based plant mortality model that incorporates individual based measurements of growth as well as the species trait wood density. Furthermore, we extend existing individual-based approaches \citep[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship (hereon in referred to as growth-dependent hazard). We parameterise this model using 15-years of individual dbh growth and mortality data from 203 tropical rainforest species encompassing \Sexpr{n_inds} individuals. Using this model,coupled with cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to the null model or growth-dependent only model 2) Does allowing both hazards to vary by species improve predictive capacity? 3) Does wood density predominately affect baseline hazards, or does it affect growth-dependent hazards or both? 4) What better predicts plant mortality, absolute dbh growth or basal area growth?

@@

\subsection{Data}
We parameterise plant mortality models using individual growth and survival data collected from a 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\degree;N, 79.85\degree;W). The plot is predominately undisturbed old-growth rainforest. The climate is warm throughout the year with a mean annual temperature of 26.1\degree;. Rainfall is seasonal with most of the 2500 mm falling between April and November. The island is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI) and the permanent plot was established in 1980. Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in (REFS).

Within the 50-ha BCI plot the diameter at breast height and survival status of all free-standing woody plants at least 1.3 m tall with a diameter greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter \citep{Condit:2012nz}. For the purpose of modelling mortality as a function of growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre. We also excluded individuals based on several rules: 1) those that did not survive at least two censuses (two being required to estimate growth rates); 2) those that were not consistently measured at 1.3 m above ground; 3) those that were multi-stemmed; or 4) those that resprouted or '\textit{returned from the dead}'. We also discarded species that do not exhibit secondary growth (e.g. palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density. Extreme outliers: stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial dbh were also removed. Using this dataset, we then randomly selected one census observation per individual. This was done in order to reduce computational requirements. In total, \Sexpr{n_inds} individuals of \Sexpr{n_spp} species were included in our analysis.

@@

\subsection{Estimating true growth}
Field measurements almost always include observation error, and these errors can potentially manifest into biologically unrealistic scenarios that may ultimately bias interptations. In our dataset, \Sexpr{proportion_negative}\% of the observations were of negative dbh growth. While it is known that dbh can decline due to changes in water availability, growth We consider a biologically unrealistic scenario

Here we quantify observation error by using another BCI dataset explicitly collected for this purpose. This dataset consistend of \Sexpr{n_obs_errors} randomly selected trees remasured within 30 days (see XXXX). By assuming that no growth occurred between the two observations, and that observation error did not vary with size, the standard deviation of measurement discrepencies

explicitly collected to examine  consisting of \Sexpr{n_obs_errors} randomly selected trees remeasured within 30 days (see XXXX). By assuming that no growth occurred between the two observations, we estimated the standard deviation of measurement error 
$\sigma_{error}$ was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within +/- 1.47 cm of the true diameter. We then used this standard deviation of 0.75 in another Bayesian model to estimate the initial and final true dbh (i.e. observation - measurement error) for each individual. This was done by first modelling the observed initial ($dbh_{t-1}$) and final dbh ($dbh_{t}$) of each individual, $i$, as random realisations from a normal distribution centered on their respective true estimates (i.e. $\widehat{dbh1_i}$ and $\widehat{dbh2_i}$), with a standard deviation of 0.75 (i.e. $\sigma_{error}$).

\bibliography{refs}

### STILL CLEANING
# Methods

## Data

We parameterise our models using individual growth and survival data collected from a 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15&deg;N, 79.85&deg;W). The plot is predominately undisturbed old-growth rainforest. The climate at Barro Colorado Island is warm throughout the year with a mean annual temperature of 26.1&deg;. Rainfall is seasonal with most of the 2500 mm falling between April and November (REF). The island is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI), which was granted long-term custodianship over the island by Panama's Environmental Authority. STRI establish the 50-ha plot as a permanent census in 1980. Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in (REFS).

Within the 50-ha BCI plot the diameter at breast height and survival status of all free-standing woody plants with a diameter greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter [@Condit:2012nz]. For the purpose of this study, we discarded data collected in 1982 and 1985 censuses because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre. We excluded individual trees that did not survive at least two censuses (two being required to estimate growth rates), were not consistently measured at 1.3 m above ground, were multi-stemmed, or had resprouted or 'returned from the dead'. We also discarded species that do not exhibit secondary growth (e.g. palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density. Extreme outliers: stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial dbh were also removed. Using this dataset, we then randomly selected one observation per individual. This was done in order to reduce computational requirements. In total, 180,500 individuals of 203 species were included in our analysis.

Wood density for each species was estimated by coring trees located within 15 km of the BCI plot because increment borers are prohibited within the plot [@Wright:2010fl]. Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100&deg;C) and dividing by the fresh volume (as measured by water displacement). Wood density was then calculated as an area-weighted average (g cm$^-3$), where area refers to the annulus represented by each piece, assuming a circular trunk.

## Measurement error and calculating true growth

Observed dbh measurements are likely to include measurement error that can manifest into biologically unrealistic scenarios of negative growth and ultimately erroneous conclusions. We estimate this measurement error using another BCI dataset consisting of 1562 randomly selected trees remeasured within 30 days (see XXXX). By assuming that no growth occurred between the two observations, we estimated the standard deviation of measurement error 
$\sigma_{error}$ was estimated to be 0.75 (sd = 0.01), meaning that 95\% of the measurement errors were within +/- 1.47 cm of the true diameter. We then used this standard deviation of 0.75 in another Bayesian model to estimate the initial and final true dbh (i.e. observation - measurement error) for each individual. This was done by first modelling the observed initial ($dbh_{t-1}$) and final dbh ($dbh_{t}$) of each individual, $i$, as random realisations from a normal distribution centered on their respective true estimates (i.e. $\widehat{dbh1_i}$ and $\widehat{dbh2_i}$), with a standard deviation of 0.75 (i.e. $\sigma_{error}$).

\begin{equation} \label{eq:obs dbh1}
dbh_{i,t-1} \sim \textrm{N}(\widehat{dbh_{i,t-1}}, 0.75). \end{equation}

\begin{equation} \label{eq:obs dbh2}
dbh_{i,t} \sim \textrm{N}(\widehat{dbh_{i,t}}, 0.75). \end{equation}

$\widehat{dbh_{i,t-1}}$ was estimated from a lognormal distribution:

\begin{equation} \label{eq:true dbh1}
\widehat{dbh_{i,t-1}} \sim \textrm{lognormal}(\mu_{dbh_{i,t-1}}, \sigma_{dbh_{t-1}}), \end{equation}

where, $\mu_{dbh_{t-1}}$ defines the log of the grand mean initial dbh, and $\sigma_{dbh_{t-1}}$ defines that variation around that mean. $\mu_{dbh_{t-1}}$ was estimated using a normal prior $\text{N}(0, 2.5)$, which represents our prior expectation that the mean initial dbh would likely occur between 0 and 148 cm. $\sigma_{dbh{t-1}}$ was estimated using a weakly informative half cauchy prior centered on zero and a scale parameter of 2.5.

$\widehat{dbh_{i,t}}$ was estimated as:

\begin{equation} \label{eq:true dbh2}
\widehat{dbh_{i,t-1}} + R_i \, T_i, 
\end{equation}

where, $R_i$ represents the true annual increase in dbh (cm) and $T_i$ represents the number of days between measurements divided by 365.25. We modelled $R_i$ as a random realisation from a half normal distribution centered on zero and a standard deviation of 5, conveying our prior expectation that annual increases in dbh would be between 0 and 10 cm/yr. We used a half normal as it ensures dbh growth cannot be negative.

\begin{equation} \label{eq:obs dbh increment}
R_i \sim \textrm{N+}(0, 5). 
\end{equation}

For computational efficiency, we used the posterior means of $dbh_{t-1}$, $dbh_t$ to calculate:

dbh growth rate:  $\frac{\widehat{dbh_{i,t}} - \widehat{dbh_{i,t-1}}}{T_i}$ 
basal area growth rate: $\frac{(\pi \, 0.25 \,\widehat{dbh_{i,t}}) - (\pi \, 0.25 \, \widehat{dbh_{i,t-1}})}{T_i}$.

These estimates of individual post growth were then used in the subsequent mortality analysis.

## Mortality model description

We extend a common functional form used to model the growth-mortality relationship [e.g. @Kobe:1995tw] to include a growth-independent baseline hazard, such that the functional form becomes
\begin{equation} \label{eq:lambda}
\lambda = \overbrace{\alpha \, \exp\left(-\beta \, G \right)}^{\text{Growth dependent}} + \overbrace{\gamma}^{\text{Growth independent}}
\end{equation}

By adding $\gamma$ we allow the functional form to asymptote at a mortality rate above zero, which allows for deaths to occur at high growth rates that would otherwise not be accounted for. The advantage of this functional form is that all parameters are biologically interpretable. $\alpha$ represents the intercept and biologically defines the sum of mortality rate at zero growth and baseline hazard. $\beta$ represents the function's decay, or in other words, the sensitivity of mortality rates to change with growth rate. $\gamma$, as stated before, represents asymptote and is referred to as the baseline hazard. $\alpha$ and $\beta$ are intrinsically linked to growth rate, and as such are growth-dependent parameters. By contrast, $\gamma$ accounts for mortality not related to growth caused by stachastic events (e.g. windfall, fire) and consequently is a growth-independent parameter.

To incorporate varability in species mortality rates, we add allowed $\alpha$, $\beta$ and $\gamma$ to vary by wood density - a species trait thought to encapsulate a species growth strategy (REFs). In addition to this, we also allowed the three parameters to vary by species ID, by doing this, we effectively allow the model to partition variation explained by wood density from the residual variation explained by species and individual levels [@Gelman:2007te; @Kery:2010tp].

## Bayesian Hierarchical Model implementation

We have binary observations $y_i$ (0 = alive, 1 = dead) for each individual $i$. We also have a hypothesised underlying mortality model giving an instantaneous rate of mortality (or hazard) per unit time (eq. \ref{eq:lambda}). Based on standard demographic theory, the probability of death for individual $i$ across the census interval $t_1\rightarrow t_2$ is
\begin{equation} \label{eq:pred_model}
p_{i,t_1\rightarrow t_2} = 1 - \exp\left(\Lambda_{i,t_1\rightarrow t_2} \right),
\end{equation}
where $\Lambda_{i,t} = \int_{t1}^{t2} \lambda(t) \, dt$ is the cumulative hazard for that individual. Given estimates of $\lambda$, one can model the observations as a random realization from a Bernoulli distribution based on the probability that individual $i$ would die, $p_{i,t_1\rightarrow t_2}$:
\begin{equation} \label{eq:obs_model}
y_{i,t_1\rightarrow t_2} \sim \textrm{bernoulli}(p_{i,t_1\rightarrow t_2}).\end{equation}

Our first hypothesis is that the cumulative hazard rate within a census interval is reasonably approximated by the true growth rate (estimated as above) in the previous interval
\begin{equation} \label{eq:hazard2}
\Lambda_{i,t} \approx (t_2-t_1) \lambda (t_1).
\end{equation}

Second, we assume that the three parameters of eq. \ref{eq:lambda} are influenced by 3 factors: wood density, species ID and census year. Because wood density is measured at species-level, its effect is logically a subset of the overall species level effect.

To account for species strategy and partition variance into explained and unexplained at the observation and species level we modelled $\alpha$, $\beta$ and $\gamma$ via the following submodels:
\begin{equation} \label{eq:submodels}
\alpha_s = \alpha_{0,s} \, \rho_s^{\alpha_{1}},\\
\beta_s = \beta_{0,s} \, \rho_s^{\alpha_{1}}, \\
\gamma_s = \gamma_{0,s} \, \rho_s^{\alpha_{1}}. 
\end{equation}

Here $\alpha_{0,s}$,$\beta_{0,s}$,$\gamma_{0,s}$ represents the species-level random effect of species $s$ within which individual $i$ belongs and $\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of standardised wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively. The form of submodels in eq \ref{eq:submodels} ensures that parameters remain positive. On a log scale these submodels \ref{eq:submodels} become additive (e.g. $\log \alpha_s = \log \alpha_{0,s} + \alpha_{1} \, \log \rho_s$, equivalent to a classic proportional hazards model widely used in survival analysis. We standardised wood density via the transformation: $\rho_s = \frac{\rho}{0.6}$), an operation which is equivalent to centering a log-transformed variable at a density of 0.6 g cm${^-3}$. This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant) [@Gelman:2007te].

The species random effects on each parameter were assumed to be a random realistation from a log-normal distribution:
\begin{equation} \label{eq:RE_species}
\alpha_{0,s} \sim \text{log-normal}\left(\mu_{\log \alpha_0}, \sigma_{\log \alpha_0}\right),\\
\beta_{0,s} \sim \text{log-normal}\left(\mu_{\log \beta_0}, \sigma_{\log \beta_0}\right),\\
\alpha_{0,s} \sim \text{log-normal}\left(\mu_{\log \gamma_0}, \sigma_{\log \gamma_0}\right),\\
\end{equation}

We estimated $\mu_{\log \alpha_0}$, the average log species effect at zero growth using the following informative prior:

\begin{equation} \label{eq:alpha0}
\mu_{\log \alpha_{0}} \sim \text{normal}\left(-0.61, 0.57\right).\\
\end{equation}

This distribution was derived from data published by Kobe et al 1995, who estimated the probability of dying at zero growth/2.5 years for juveniles of 10 hardwood species across north-eastern America. Note that these data are also used to define the growth-mortality relationship in SORTIE-ND a spatially explicit forest dynamic model (REF). We converted these probabilities/2.5 years to instantaneous hazard rates by using $\lambda = -\log(1 - Pr(Death))/2.5$, and then log transforming these values and calculating the mean (i.e. -0.61) and standard error (i.e. 0.57). Based on this prior, we assumed that the average probability of death was between 16 and 81% with the central tendency being 42%/yr.

Similarly, we used prior information from Korning and Balslev et al (1994) to estimate, $\mu_{\log \gamma_0}$, the average log species effect on $\gamma$, such that:

\begin{equation} \label{eq:gamma0}
\mu_{\log \gamma_0} \sim \textrm{N}\left(-3.86, 0.2\right).\\
\end{equation}

>>>>

\documentclass[a4paper,12pt]{article}
\usepackage[osf]{mathpazo}
\usepackage{ms}
\usepackage[]{natbib}
\usepackage{gensymb}
\usepackage{amsmath}
\usepackage[title, titletoc, toc]{appendix}
%\raggedright

@@

@@

<<preload, echo=FALSE>>=
n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique)*100)
n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
@

@@

Wood density, the biomass invested per unit wood volume, is one functional trait that has been consistently shown to be negatively correlated with tropical tree mortality \citep{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}. The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to random events such as stem breakage \citep{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \citep{Hacke:2001kj, Jacobsen:2005fx} and fungal and pathogen attack \citep{Augspurger:1984wx}. As such, it is hypothesized that species with higher wood density are likely to exhibit lower baseline hazards. However, research has also shown that wood density may increase tolerance to low growth rates, and thus alter the growth-mortality relationship. For example, species with high wood density are more tolerant shade \citep{vanGelder:2006exa} and competition \citep{Kunstler:2016km}. Studies have also shown that wood density may reduce drought induced carbon starvation \citep{McDowell:2011dr} and the susceptibility of ‘stressed’ individuals to opportunistic invertebrate, pathogen or fungal attack \citep{Augspurger:1984wx, McDowell:2011dr}.

@@

In this study, we reconcile both intraspecific and interspecific theory by developing an individual based models of plant mortality that incorporate individual based measurements of growth as well as the species trait wood density. Furthermore, we extend existing individual-based approaches \citep[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship (hereon in referred to as growth-dependent hazard). We parameterise this model using 15-years of individual dbh growth and mortality data from 203 tropical rainforest species encompassing \Sexpr{n_inds} individuals. Using this model,coupled with cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to the null model or growth-dependent only model 2) Does allowing both hazards to vary by species improve predictive capacity? 3) Does wood density predominately affect baseline hazards, or does it affect growth-dependent hazards or both? 4) What better predicts plant mortality, absolute dbh growth or basal area growth?

@@

\subsection{Data}
We parameterise plant mortality models using individual growth and survival data \citep{Condit:2012nz} collected from a relatively undisturbed 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\degree;N, 79.85\degree;W). The climate on the island is warm throughout the year with a mean annual temperature of 26.1\degree and rainfall is seasonal with most of the 2500 mm falling between April and November. BCI is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI) and the permanent plot was established in 1980. Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in .

Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and contained a diameter at breast height greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter \citep{Condit:2012nz}. For the purpose of modelling mortality as a function of growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre. We also excluded individuals based on several rules: 1) those that did not survive at least two censuses (two being required to estimate growth rates); 2) those that were not consistently measured at 1.3 m above ground; 3) those that were multi-stemmed; or 4) those that resprouted or '\textit{returned from the dead}'. We also discarded species that do not exhibit secondary growth (e.g. palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density. Extreme outliers: stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial dbh were also removed. In order to reduce computational requirements we then subsampled this dataset by randomly selecting one census observation per individual. In total, \Sexpr{n_inds} individuals of \Sexpr{n_spp} species were included in our analysis.

@@

\subsection{Estimating true growth}

Field measurements almost always include observation error, errors that may ultimately bias conclusions. In our dataset, \Sexpr{proportion_negative}\% of observed growth rates were negative. While there are physiological processes (e.g. water depletation and bark drop) that can lead to trunk shrinkages overtime \citep{Pastur:2007kl}, it is biologically impossible for growth (i.e. the creation of new biomass) to be negative \citep{Clark:2007ve}. To ensure our model was not biased by these implausable values we  estimated '\textit{true growth}'(i.e. observed growth minus observation error) by first using another BCI dataset collected explicitly to examine obervation errors \citep{Condit:2012nz}. This dataset consisted of \Sexpr{n_obs_errors} randomly selected trees that had their dbh measured twice within a 30 day period \citep{Condit:2012nz}. Assuming no growth occurred over this period, any discrepency in measurements can be attributable to observation error. The standard deviation of these discrepencies was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within +/- 1.47 cm of the true diameter. Having an estimate of observation error, we then built an empirical Bayesian model to estimate '\textit{true}' initial and final dbh, and subsequently,'\textit{true growth}' for each individual. We did this by first modelling the observed initial ($dbh_{t-1}$) and final dbh ($dbh_{t}$) of each individual, $i$, as random realisations from a normal distribution centered on their respective true estimates (i.e. $\widehat{dbh1_i}$ and $\widehat{dbh2_i}$), with the standard deviation fixed at \Sexpr{obs_error_sigma} (i.e. the observed measurement error):

\begin{equation}\label{eq:true_growth_obs} 
\begin{aligned}
dbh_{i,t-1} &\sim \mathcal{N}(\widehat{dbh}_{i,t-1}, 0.75) \\
dbh_{i,t} &\sim \mathcal{N}(\widehat{dbh}_{i,t}, 0.75)
\end{aligned} 
\end{equation}

The initial dbh, $\widehat{dbh}_{i,t-1}$, was sampled as a random realisations from a lognormal distribution:
\begin{equation} \label{eq:true_dbh1}
\widehat{dbh}_{i,t-1} \sim \mathcal{LN}(\mu_{dbh_{i,t-1}}, \sigma_{dbh_{t-1}}), 
\end{equation}
where, $\mu_{dbh_{t-1}}$ is the expected mean of log initial dbh, and $\sigma_{dbh_{t-1}}$ defines that variation around that mean. $\mu_{dbh_{t-1}}$ was estimated using a normal prior $\mathcal{N}(0, 2.5)$, which represents our prior expectation that the mean initial dbh would be greater than zero but less than 150 cm. $\sigma_{dbh{t-1}}$ was estimated using a weakly informative half cauchy prior centered on zero and a scale parameter of 2.5.

The final dbh measurement, $\widehat{dbh}_{i,t}$ was estimated as:
\begin{equation}\label{eq:true_dbh2}
\widehat{dbh}_{i,t} = \widehat{dbh}_{i,t-1} + R_i \, T_i, 
\end{equation}
where, $R_i$ represents the annual true increase in dbh (cm) and $T_i$ represents the number of days between measurements divided by 365.25. We modelled $R_i$ as a random realisation from a lognormal distribution centered on zero and a standard deviation of 1. This prior conveyed our expectation that annual increases in dbh were unlikely to be greater than 7 cm/yr. Sampling from a lognormal distribution also ensured growth rates could not be negative.

Using this model we then used \stan's \textit{optimizing} function to maximise the joint posterior and extract point estimates of true initial dbh, $\widehat{dbh}_{i,t-1}$, true final dbh, $\widehat{dbh}_{i,t}$, and true annual dbh growth $R_i$. We used annual diameter growth, $R_i$, and annual basal area growth:$\frac{(\pi \, 0.25 \,\widehat{dbh}_{i,t}) - (\pi \, 0.25 \, \widehat{dbh}_{i,t-1})}{T_i}$ in the subsequent mortality analysis.

\subsection{Mortality analysis}

In survival analysis, the probability that an individual dies between $t_1$ and $t_2$, $p_{t_1\rightarrow t_2}$ is defined by:
\begin{equation} \label{eq:mortality_model}
p_{t_1\rightarrow t_2} = 1 - \exp\left((t_2-t_1)\,\lambda_i\right)
\end{equation}
where, $\lambda$, is the instantaneous hazard rate of death per unit time \citep{Kleinbaum:2012tn}.

Here we examine the performance of three functional forms of $\lambda$ to predict observed patterns of tropical tree mortality. The simplest model (eqn: \ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plants growth rate. The second model (eqn: \ref{growth_haz} describes a growth-dependent hazard which assumes plants can only die due to a growth related cause. In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached. This negative exponential relationship has been used in other studies to model growth-dependent mortality \citep[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \citep{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}. The third model (eqn: \ref{add_haz}, and potentially the most biologically realistic, is the summation of the two previous models. Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: that due to growth and that due to chance (i.e. stochastic events such as windthrow and fire). The parameters of eqn: \ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept (i.e. sum of both growth dependent and growth independent mortality at zero growth rate) and biologically defines the instantaneous mortality hazard $\lambda$ at zero growth. $\beta$ represents the exponential decay rate or the sensitivity of hazard rates to change with growth rate; and $\gamma$, is the asymptote and defines the baseline hazard. $\alpha$ and $\beta$ are intrinsically linked to growth rate, and as such are growth-dependent parameters. By contrast, $\gamma$ is a growth-independent parameter that accounts for stochastic events (e.g. windfall, fire) that can kill a plant irrespective of its growth rate.
\begin{align}
\lambda &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
\lambda &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
\lambda &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} + 
\overbrace{\gamma}^{\text{Baseline hazard}} \label{add_haz}
\end{align}

A central aim of this study is to identify what is the most predictive parameterisation of $\lambda$. To do this, we used K-fold cross-validation and four iterative stages of model fitting and comparison. Stage 1 involved comparing predictive performance of the three $\lambda$ models described above (eqn: \ref{base_haz,growth_haz,add_haz}). In this stage each model was fitted with fixed parameters (i.e. parameters were constant across species) and then again by allowing parameters to varying by species. This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) the extent to which species hazard rates differ. In stage 2 we used the most predictive model from stage 1 to examine whether dbh growth or basal area growth was a better predictor of plant mortality. In stage 3 we used the best model from stage 2, but without species random effects, to determine what combination of parameter-wood density effects best predicted tree mortality. This stage also allowed us to determine whether wood density primarily affected growth-dependent or growth-independent (i.e. baseline) mortality. Finally, in stage 4, we fit the best model based on the previous three steps to the entire dataset with species random effects on all parameters. We then examined wood density effect sizes and the proportion of species variability that it explained.

\subsection{Model fitting}

Our observations are of survival status $y_i$ (0 = alive, 1 = died) for each individual $i$ between two censuses. We model these observations as random realizations from a Bernoulli distribution based on the probability that individual $i$ dies between $t_1$ and $t_2$, $p_{i,t_1\rightarrow t_2}$:
\begin{equation}
y_{i,t_1\rightarrow t_2} \sim \textrm{Bernoulli}(p_{i,t_1\rightarrow t_2}).
\end{equation}

We modelled $p_{i,t_1\rightarrow t_2}$ using a complementary log-log link as:
\begin{equation}
\log(-\log(1 - p_{i,t_1\rightarrow t_2})) = \log((t_2-t_1)\,\lambda_i)
\end{equation}.
This parameterisation allowed for additive hazards such as those in \ref{add_haz} and is equivilant to a competiting risks analysis \citep{Kleinbaum:2012tn}.

As we have multiple hypothesized functional forms of $\lambda$ (e.g. eqn: X ), each containing parameters that may or may not vary by species, we outline the fitting procedure of our most complicated model of which all others are nested within:
\begin{equation}
\lambda_i = \overbrace{\alpha_{s[i]} \, \exp\left(-\beta_{s[i]} \, Growth_i \right)}^{\text{Growth dependent}} + 
\overbrace{\gamma_{s[i]}}^{\text{Growth independent}}.
\end{equation}

In this model, $\alpha_s$, $\beta_s$ and $\gamma_s$ varying by species a function of both wood density (measured at species level) and a species random effect:
\begin{equation}\label{eq:submodels}
\begin{aligned}
\alpha_s &= \alpha_{0,s} \, \rho_s^{\alpha_{1}},\\
\beta_s &= \beta_{0,s} \, \rho_s^{\alpha_{1}}, \\
\gamma_s &= \gamma_{0,s} \, \rho_s^{\alpha_{1}}. 
\end{aligned}
\end{equation}
Here $\alpha_{0,s}$,$\beta_{0,s}$,$\gamma_{0,s}$ represents the species-level residual error not explained by wood density for species $s$ within which individual $i$ belongs. $\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of standardised wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively. The form of submodels in eqn: \ref{eq:submodels} ensures that parameters remain positive. On a log scale these submodels become additive (e.g. $\log \alpha_s = \log \alpha_{0,s} + \alpha_{1} \, \log \rho_s$. We standardised wood density via the transformation: $\rho_s = \frac{\rho}{0.6}$), an operation which is equivalent to centering a log-transformed $\rho$ at a density of 0.6 g cm${^-3}$. This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6 cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant).

Species random effects were modelled as random realisations from log-normal distributions:
\begin{equation}
\begin{aligned}
\alpha_{0,s} &\sim \mathcal{LN}\left(\mu_{\log \alpha_0}, \sigma_{\log \alpha_0}\right),\\
\beta_{0,s} &\sim \mathcal{LN}\left(\mu_{\log \beta_0}, \sigma_{\log \beta_0}\right),\\
\gamma_{0,s} &\sim \mathcal{LN}\left(\mu_{\log \gamma_0}, \sigma_{\log \gamma_0}\right).
\end{aligned}
\end{equation}

Both $\mu_{\log \alpha_0}$, the average log species residual error at zero growth, and $\mu_{\log \gamma_0}$, the average baseline log species residual error, from normal distributions centered on zero with a standard deviation of 2.5.
\begin{equation}
\mu_{\log \alpha_{0}},\mu_{\log \gamma_0} \sim \mathcal{N}\left(0, 2.5\right).\\
\end{equation}

We estimated $\mu_{\log \beta_0}$, the average species error on the exponential decay rate, using a weak normal prior centered on zero with a standard deviation of 5. 
\begin{equation}
\mu_{\log \beta_0} \sim \mathcal{N}\left(0, 5\right).\\
\end{equation}
As we were fitting these models to multiple growth measures (i.e. dbh increment, basal area growth), we intentially made this prior weak, such that the range of possible values spanned from a rapid exponential decay to no change in mortality with increasing growth.

Standard deviations associated with species random effects (i.e. $\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$), were estimated using positive half cauchy priors centered on zero with a scale of 2.5.
\begin{equation}
\sigma_{a0},\sigma_{b0},\sigma_{c0} \sim \textrm{Cauchy+}\left(0, 2.5\right).\\
\end{equation}
For models that did not contain species random effects, the above sigma's were not calculated and only the hyperparameter means (i.e. $\mu_{\log \alpha_0}$,$\mu_{\log \beta}$,$\mu_{\log \gamma_0}$) were estimated.

Wood density effects (i.e. $\alpha_1$, $\beta_1$ and $\gamma_1$) were estimated using normal priors centered on zero and a standard deviation of 5:
\begin{equation} \label{eq:rho_effects}
\alpha_1, \beta_1, \gamma_1 \sim \mathcal{N}\left(0, 5\right).\\
\end{equation}

All models were fit using package rstan version 2.8.1 \citep{StanDevelopmentTeam:2015uz}. Three independent chains were executed and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \citep{Gelman:1992ts, Brooks:1998ju}. After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.

\subsection{Model evaluation}

We examined predictive accuracy of each model by using K-fold cross-validation. This involved randomly partitioning the data into ten folds of X observations, ensuring that each K-fold had at least one observation per species. We then fitted models to nine folds and predicted to the tenth fold. This was repeated until each fold had been used nine times for model fitting, and once as a heldout dataset for predicting. Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy (Stone:1997tg).
